An elementary and constructive proof of Grothendieck's generic freeness lemma

### 2018-06-02 paper-generic-freeness.tex
Incorporated the nontrivial suggestion by Rosolini

\documentclass{amsart}

@@

@@

\newcommand{\XXX}[1]{\textbf{XXX: #1}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\defeq}{\vcentcolon=}
\DeclareMathOperator{\Spec}{Spec}

@@

\begin{thm}\label{thm:algebraic}Let~$A$ be a reduced ring. Let~$B$ be
an~$A$-algebra of finite type. Let~$M$ be a finitely generated~$B$-module.
If~$f = 0$ is the only element of~$A$ such that
\begin{enumerate}
\item the~$A[f^{-1}]$-modules $B[f^{-1}]$ and $M[f^{-1}]$ are free,
\item the~$A[f^{-1}]$-algebra~$B[f^{-1}]$ is of finite presentation and
\item the~$B[f^{-1}]$-module~$M[f^{-1}]$ is finitely presented,
\end{enumerate}
then~$1 = 0$ in~$A$.
\end{thm}

Previously known proofs either only cover the case where~$A$ is a Noetherian
integral domain, where one can argue by \emph{dévissage} (see for
instance~\cite[Lemme~6.9.2]{ega-4-2},
\cite[Thm.~24.1]{matsumura:commutative-ring-theory} or
\cite[Thm.~14.4]{eisenbud:commutative-algebra}), or proceed in a series of
intermediate steps, reducing to that case (see for
instance~\cite{staats:generic-freeness} or~\stacksproject{051Q}); but in fact,
a direct proof is possible and shorter. The new proof unveils a certain
combinatorial aspect to Grothendieck's generic freeness lemma, does not require
any advanced prerequisites in commutative algebra and does not use the axiom of
choice or the law of excluded middle. It is purely element-based, not referring
to ideals of~$A$.
\enlargethispage{3em}

@@

We remark that the proof takes a somewhat curious course: Our goal is to
verify~$1 = 0$, but as an intermediate step we verify that~$M$ is free, which
after the fact will be a trivial statement. The general proof in the next
section will have a similar style. This approach is reminiscient of
Richman's uses of trivial rings~\cite{richman:trivial-rings}.
\enlargethispage{1em}

\begin{figure}[t]
  \captionsetup{width=.9\linewidth}
  \begin{tikzpicture}[scale=0.9]
    \draw[step=1cm,gray,very thin] (0,0) grid (8,8);

@@

We show that~$(w_J)_{J \in \J'}$ is a basis of~$M$ by verifying linear
independence. Thus let~$\sum_J a_J w_J = 0$ in~$M$. We show that all
coefficients in this sum are zero, starting with the largest
appearing index~$J$: In the module~$M[a_J^{-1}]$ over the localized ring~$A[a_J^{-1}]$,
the vector~$w_J$ is a linear combination of generators with
smaller index. Removing~$w_J = x_1^{i_1} \cdots x_n^{i_n} v_\ell$ and also all
vectors~$x_1^{k_1} \cdots x_n^{k_n} v_\ell$ where~$k_1 \geq i_1, \ldots,
k_n \geq i_n$, we obtain a subfamily which is still good for
the localized module. The induction hypothesis, applied
to~$A[a_J^{-1}]$ and its module~$M[a_J^{-1}]$, therefore implies that $A[a_J^{-1}] = 0$.
Thus $a_J = 0$ since~$A$ is reduced.
\enlargethispage{2em}

@@

% History

% Application

>>>>

\documentclass[oneside]{amsart}

@@

\usepackage{geometry}

@@

\newcommand{\XXX}[1]{\textbf{XXX: #1}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\B}{\mathcal{B}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\defeq}{\vcentcolon=}
\DeclareMathOperator{\Spec}{Spec}

@@

\begin{thm}\label{thm:algebraic}Let~$A$ be a reduced ring (commutative, with unit). Let~$B$ be
an~$A$-algebra of finite type. Let~$M$ be a finitely generated~$B$-module.
If~$f = 0$ is the only element of~$A$ such that
\begin{enumerate}
\item the~$A[f^{-1}]$-modules $B[f^{-1}]$ and $M[f^{-1}]$ are free,
\item the~$A[f^{-1}]$-algebra~$B[f^{-1}]$ is of finite presentation and
\item the~$B[f^{-1}]$-module~$M[f^{-1}]$ is finitely presented,
\end{enumerate}
then~$1 = 0$ in~$A$.
\end{thm}

Previously known proofs either only cover the case where~$A$ is a Noetherian
integral domain, where one can argue by \emph{dévissage} (see for
instance~\cite[Lemme~6.9.2]{ega-4-2},
\cite[Thm.~24.1]{matsumura:commutative-ring-theory} or
\cite[Thm.~14.4]{eisenbud:commutative-algebra}), or proceed in a series of
intermediate steps, reducing to that case (see for
instance~\cite{staats:generic-freeness} or~\stacksproject{051Q}); but in fact,
a direct proof is possible and shorter. The new proof unveils a certain
combinatorial aspect to Grothendieck's generic freeness lemma, does not require
any advanced prerequisites in commutative algebra and does not use the axiom of
choice or the law of excluded middle. It is purely element-based, not referring
to ideals of~$A$, and doesn't use Noether normalization.

@@

We remark that the proof takes a somewhat curious course: Our goal is to
verify~$1 = 0$, but as an intermediate step we verify that~$M$ is free, which
after the fact will be a trivial statement. The general proof in the next
section will have a similar style. This approach is reminiscient of
Richman's uses of trivial rings~\cite{richman:trivial-rings}.
\enlargethispage{8em}

\begin{figure}[H]
  \captionsetup{width=.9\linewidth}
  \begin{tikzpicture}[scale=0.9]
    \draw[step=1cm,gray,very thin] (0,0) grid (8,8);

@@

We show that~$(w_J)_{J \in \J'}$ is a basis of~$M$ by verifying linear
independence. Thus let~$\sum_J a_J w_J = 0$ in~$M$. We show that all
coefficients in this sum are zero, starting with the largest
appearing index~$J$: In the module~$M[a_J^{-1}]$ over the localized ring~$A[a_J^{-1}]$,
the vector~$w_J$ is a linear combination of generators with
smaller index. Removing~$w_J = x_1^{i_1} \cdots x_n^{i_n} v_\ell$ and also all
vectors~$x_1^{k_1} \cdots x_n^{k_n} v_\ell$ where~$k_1 \geq i_1, \ldots,
k_n \geq i_n$, we obtain a subfamily which is still good for
the localized module. The induction hypothesis, applied
to~$A[a_J^{-1}]$ and its module~$M[a_J^{-1}]$, therefore implies that $A[a_J^{-1}] = 0$.
Thus $a_J = 0$ since~$A$ is reduced.

@@

Commutative algebra abounds with techniques which allow us to reduce quite
general situations to easier ones. These techniques often yield short and slick
proofs; however, they come at an expense: They are typically nonconstructive in
nature, employing for instance the axiom of choice, and do not argue using only
the data at hand, but using additional auxiliary objects such as maximal
ideals. We feel that once a subject is better understood, it is desirable to
have more informative, direct proofs available which illuminate the
statement more clearly; similar as to how bijective proofs are preferred over
calculational inductive ones in combinatorics.

Let us consider as a specific example the statement that the existence of a linear
surjection~$A^n \to A^m$ with~$n < m$ between finite free modules over an
arbitrary ring~$A$ implies~$1 = 0 \in A$. The standard proof of this fact proceeds
by contradiction and passes to the quotient~$A/\mmm$, where~$\mmm$ is a maximal
ideal of~$A$, thereby reducing to the situation that the ring is a field. In
contrast, a direct proof such as Richman's~\cite{richman:trivial-rings}
refers only to objects mentioned in the statement itself and
explicitly tells us how to deduce the equation~$1 = 0$ from the~$m$ equations
which express that each basis vector of~$A^m$ has a preimage.

In a similar fashion, the new proof of Grothendieck's generic freeness lemma
explicitly tells us how to deduce~$1 = 0$ from the given conditional
equations expressing that~$f = 0$ is the only element with properties~(1),~(2)
and~(3). The history of Grothendieck's generic freeness lemma goes back more
than fifty years; we are slightly surprised that a direct proof was
discovered only now.

Direct proofs sometimes generalize to new situations where the reduction techniques
employed in more abstract proofs cannot be applied. This is the case for
Grothendieck's generic freeness lemma, which allows for the following
generalization:

\begin{thm}\label{thm:general-generic-freeness}
Let~$(X,\O_X)$ be a ringed space (or ringed locale, or ringed topos), such
that for every local section~$s$ of~$\O_X$, if the only open on which~$s$ is
invertible is the empty one, then~$s = 0$. Let~$\B$ be a sheaf
of~$\O_X$-algebras of finite type. Let~$\E$ be a sheaf of~$\B$-modules of finite type. Then there is a dense
open~$U$ such that over~$U$,
\begin{enumerate}
\item[(a)] $\B$ and~$\E$ are locally free as sheaves of~$\O_X$-modules,
\item[(b)] $\B$ is of finite presentation as a sheaf of~$\O_X$-algebras and
\item[(c)] $\E$ is finitely presented as a sheaf of~$\B$-modules.
\end{enumerate}
\end{thm}

\begin{proof}The proof of Theorem~\ref{thm:algebraic} can be easily adapted to
this more general setting. Where that proof concludes~$a_J = 0$ for ring
elements~$a_J$ by considering the localized situation~$A[a_J^{-1}]$, we now
conclude~$a_J = 0$ for local sections~$a_J$ of~$\O_X$ by considering the
situation over the restriction to~$D(a_J)$. It is then the assumption on~$X$
instead of the reducedness assumption on~$A$ which allows us to conclude~$a_J = 0$.
We omit further details.
\end{proof}

The spectrum of a ring~$A$ is a space of the kind required by
Theorem~\ref{thm:general-generic-freeness} if and only if~$A$ is reduced,
thus Theorem~\ref{thm:general-generic-freeness} indeed
generalizes Theorem~\ref{thm:geometric}. Further examples for admissible spaces
are given by any topological, smooth or complex manifold; a corollary
of Theorem~\ref{thm:general-generic-freeness} for these examples is that
quotients of vector bundles, computed in the category of sheaves of modules,
are again vector bundles after restricting to suitable dense opens.

It is hard to say with certainty that a given proof does not generalize to a
new situation; but we do not see how this could be the case with the proofs
cited in the introduction.
